Current status and propsects of R-packages for the design of experiments

### 2021-03-05 paper.Rmd
abstract + intro

---
title: Current status and propsects of R-packages for the design of experiments
author:
  # see ?rjournal_article for more information
  - name: Emi Tanaka
    affiliation: Monash University
    address:
    - Monash University
    - Clayton campus, VIC 3800, Australia
    url: http://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email:  emi.tanaka@monash.edu
abstract: >
  An abstract of less than 150 words.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

```{r setup, cache = FALSE, include = FALSE}
library(tidyverse)
library(rvest)
library(lubridate)
library(cranlogs)
library(glue)
library(scales)
library(colorspace)
library(tidytext)
library(pluralize)
library(kableExtra)
library(igraph)
library(ggraph)

knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align='center', 
                      fig.pos='htbp')

myggtheme <- 
  theme(panel.background = element_rect(fill = NA),
        panel.grid = element_line(color = "#f6e5ee"),
        axis.text = element_text(color = "#79003e"),
        axis.line = element_line(color = "#79003e", size = 0.7),
        axis.ticks.length = unit(1.4, "mm"),
        axis.ticks = element_line(color = "#79003e", size = 0.7),
        axis.title = element_text(color = "#79003e", face = "bold"),
        strip.background = element_rect(color = "#79003e",
                                        fill = "#AD0059"),
        strip.text = element_text(color = "white"),
        plot.title.position = "plot",
        plot.title = element_text(color = "#79003e", face = "bold")) 
```

```{r cran-titles}
# Thanks to Dirk Eddelbuettel's answer on SO:
# https://stackoverflow.com/questions/11560865/list-and-description-of-all-packages-in-cran-from-within-r
url <- "http://cran.rstudio.com/web/packages/packages.rds"
db <- readRDS(url(url)) %>% 
  as.data.frame()
```

```{r}
nanalysis <- db %>% 
  filter(str_detect(tolower(Title), "analysis")) %>% 
  nrow()

ndesign <- db %>% 
  filter(str_detect(tolower(Title), "design")) %>% 
  nrow()
```

```{r doe-cran}
dat_DoE <- read_html("https://cran.r-project.org/web/views/ExperimentalDesign.html")
date_download <- Sys.Date()
cran_names <- available.packages(contriburl = "http://cran.rstudio.com/src/contrib") %>% 
  rownames() %>% 
  unique() # it should be unique
doe_pkgs <- dat_DoE %>% 
  html_nodes("li") %>% 
  html_nodes("a") %>% 
  html_text() %>% 
  .[. %in% cran_names] %>% 
  unique()

dat_survey <- read_html("https://cran.r-project.org/web/views/OfficialStatistics.html")
survey_pkgs <- dat_survey %>% 
  html_nodes("li") %>% 
  html_nodes("a") %>% 
  html_text() %>% 
  .[. %in% cran_names] %>% 
  unique()
```

@@

```{r bigram-title}
bigram_tab(doe_db, Title) %>% 
  filter(n > 3) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(caption = "The bigram of the R-package titles as provided in the DESCRIPTION file in CRAN.", 
      col.names = c("Bigram", "Count")) 
```

```{r bigram-desc}
bigram_tab(doe_db, Description) %>% 
  filter(n > 4) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(caption = "The bigram of the R-package descriptions as provided in the DESCRIPTION file in CRAN.", 
      col.names = c("Bigram", "Count"), 
      booktabs = TRUE) 
```

@@

```{r cran-data}
end <- Sys.Date() - 2 # usually 1-2 days are not available yet
start <- end - years(5) + days(2)
dldat <- cran_downloads(doe_pkgs, from = start, to = end)
```

```{r download-hist}
dldat %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>%
  ggplot(aes(total)) + 
  geom_histogram(color = "white", fill = "#AD0059") + 
  scale_x_log10(label = comma) + 
  myggtheme + 
  labs(x = glue("Total download counts from {start} to {end}"),
       y = "Number of packages") +
  scale_y_continuous(expand = c(0, 0))
```

```{r top5}
ntop <- 5

top5sum_df <- dldat %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>% 
  ungroup() %>% 
  slice_max(order_by = total, n = ntop)

top5 <- top5sum_df %>% 
  pull(package)

top5_df <- dldat %>% 
  filter(package %in% top5) %>% 
  mutate(package = fct_reorder(package, count, function(x) -sum(x))) 
```

```{r download-barplot}
top5sum_df %>% 
  mutate(package = fct_reorder(package, total)) %>% 
  ggplot(aes(total, package)) +
  geom_col(aes(fill = package)) +
  labs(x = glue("Total downloads from {start} to {end}"),
       y = "Package") + 
  scale_x_continuous(labels = comma, expand = c(0, 0)) +
  myggtheme + 
  scale_fill_discrete_qualitative(rev = TRUE) + 
  guides(fill = FALSE)
```

```{r pkg-updates, cache = TRUE, dependson="doe-cran"}
pkg_url <- "https://cran.r-project.org/web/packages/{pkg}/index.html"
pkg_archive <- "https://cran.r-project.org/src/contrib/Archive/{pkg}/"
pkg_updates <- map(top5, function(pkg) {
  last_update <- read_html(glue(pkg_url)) %>% 
    html_table() %>% 
    .[[1]] %>% 
    filter(X1=="Published:") %>% 
    pull(X2) %>% 
    ymd()

  archive_dates <- tryCatch({ 
    read_html(glue(pkg_archive)) %>% 
      html_table() %>%
      .[[1]] %>% 
      pull(`Last modified`) %>% 
      ymd_hm() %>% 
      na.omit() %>% 
      as.Date()
  }, error = function(e) {
    NULL
  })
  c(archive_dates, last_update)
})
names(pkg_updates) <- top5

updates <- unlist(pkg_updates) %>% 
  enframe("package", "update") %>% 
  # unlist converts date to integers
  mutate(update = as.Date(update, origin = "1970-01-01"),
         # need to get rid of the numbers appended to pkg names
         package = str_extract(package, paste0(top5, collapse="|")),
         package = factor(package, levels = top5)) %>% 
  filter(between(update, start, end))
```

```{r download-timeplot, dependson="top5"}
ggplot(top5_df, aes(date, count, color = package)) +
  # add shadow lines
  geom_line(data = rename(top5_df, package2 = package), 
            color = "gray", aes(group = package2)) +
  # add date when package was updated
  geom_vline(data = updates, aes(xintercept = update),
             linetype = "dashed", color = "#79003e") + 
  # the trend line
  geom_line() +
  scale_y_log10() +
  facet_grid(package ~ .) + 
  labs(title = glue("Top 5 downloaded DoE packages from {start} to {end}")) + 
  scale_color_discrete_qualitative() +
  guides(color = FALSE) +
  myggtheme
```

## Introduction

`test`
\code{test}

@@

\ctv{ExperimentalDesign}

\ctv{OfficialStatistics}

@@

```{r, Rlogo, echo=FALSE, fig.cap='The logo of R.', out.width='2in', fig.align='center', fig.pos='htbp'}
knitr::include_graphics(here::here('Rlogo.pdf'))
```

>>>>

---
title: Current status and propsects of R-packages for the design of experiments
author:
  # see ?rjournal_article for more information
  - name: Emi Tanaka
    affiliation: Monash University
    address:
    - Monash University
    - Clayton campus, VIC 3800, Australia
    url: http://emitanaka.org/
    orcid: 0000-0002-1455-259X
    email:  emi.tanaka@monash.edu
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of experiment plays a critical role in increasing the quality of experimental data.  In this article I describe the current state of the R-packages for the design of experiments through textual analysis and download trends. I discuss also the software design of widely utilised R packages in the field of experimental design and conclude with discussion of some future prospects for the field.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

```{r setup, cache = FALSE, include = FALSE}
library(tidyverse)
library(rvest)
library(lubridate)
library(cranlogs)
library(glue)
library(scales)
library(colorspace)
library(tidytext)
library(pluralize)
library(kableExtra)
library(igraph)
library(ggraph)
library(ggrepel)
library(patchwork)
library(ggwordcloud)
tocache <- TRUE

knitr::opts_chunk$set(echo = FALSE, 
                      cache = FALSE,
                      cache.path = "cache/",
                      fig.align='center', 
                      fig.pos='htbp', 
                      fig.width = 6,
                      message = FALSE,
                      warning = FALSE)

theme_set(
  theme(panel.background = element_rect(fill = NA),
        panel.grid = element_line(color = "lightgray"),
        axis.text = element_text(color = "black"),
        axis.line = element_line(color = "black", size = 0.7),
        axis.ticks.length = unit(1.4, "mm"),
        axis.ticks = element_line(color = "black", size = 0.7),
        axis.title = element_text(color = "black", face = "bold"),
        strip.background = element_rect(color = "black",
                                        fill = "black"),
        strip.text = element_text(color = "white"),
        plot.title.position = "plot",
        plot.title = element_text(color = "black", hjust = 0)))
```

```{r ctv}
doe_pkgs <- ctv:::.get_pkgs_from_ctv_or_repos("ExperimentalDesign", 
                                              repos = "http://cran.rstudio.com/")[[1]]
doe_pkgs <- c(doe_pkgs, "DeclareDesign")
npkgs <- length(doe_pkgs)
```

```{r cran-data, cache = tocache}
end <- Sys.Date() - 2 # usually 1-2 days data are not available yet
start <- end - years(2) + days(2)
dldat <- cran_downloads(doe_pkgs, from = start, to = end)
```

```{r cran-data2, cache = tocache}
start2 <- end - years(5) + days(2)
dldat2 <- cran_downloads(doe_pkgs, from = start2, to = end)
```

# Introduction

The critical role of data collection is well captured in the expression "garbage in, garbage out" -- in other words, if the collected data is rubbish then no analysis, however complex it may be, can make something out of it. Methods for data collection can be dichotomised by the type of data collected -- namely, experimental or observational -- or alternatively, categorised into experimental design or survey design.  Survey designs often, although not always, aim to collect observational data whilst experimental designs exclusively center on experimental data. This dichotomisation, to a great extent, is seen in CRAN task views where R-packages in experimental design are in \ctv{ExperimentalDesign} and R-packages in survey designs are in \ctv{OfficialStatistics}.

In the CRAN task view of \ctv{ExperimentalDesign}, there are `r npkgs` R packages for experimental design and analysis of data from experiments, henceforth referred to as "DoE packages" in this paper. The sheer quantity and variation in the output experimental design in the R-packages are arguably unmatched with any other programming languages, e.g. in Python [@python], only a handful of libraries that generate design of experiment exist (namely `pyDOE`, `pyDOE2`, `dexpy`, `experimenter` and `GPdoemd`) with limited outputs.

# Explorative Analysis

```{r download-dist, dependson = "cran-data"}
dldat_sum <- dldat %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>% 
  arrange(desc(total)) %>% 
  mutate(rank = 1:n())

dl1plot <- ggplot(dldat_sum, aes(x = "", y = total/1000)) + 
  geom_violin() +
  geom_boxplot(width = 0.1) + 
  scale_y_log10(label = comma) + 
  labs(x = "",
       y = "Download counts in the last 2 year ('000s)", title = "(A)") + 
  geom_text_repel(data = slice_max(dldat_sum, total, n = 13), 
                  nudge_x = rep(c(0.2, -0.2, 0.2, -0.2, 0.15), c(3, 2, 2, 3, 3)),
                  size = 10,
                  aes(label = package)) + 
  theme(axis.ticks.length.x = unit(0, "mm"))

dldat_sum5 <- dldat2 %>% 
  group_by(package) %>% 
  summarise(total = sum(count)) %>% 
  arrange(desc(total)) %>% 
  mutate(rank = 1:n())

top15 <- dldat_sum %>% 
  left_join(dldat_sum5, by = "package") %>% 
  slice_max(total.x + total.y, n = 15) %>% 
  select(package, rank.x, rank.y) %>% 
  pivot_longer(-package) %>% 
  mutate(name = ifelse(name=="rank.x", "Last 2 years", "Last 5 years"))

dl2plot <- ggplot(top15, aes(name, value, group = package)) +
  geom_line() + 
  geom_point() +
  scale_y_reverse(breaks = 1:15) +
  scale_x_discrete(expand = expansion(c(0.1, 0.5))) +
  labs(x = "", y = "Ranking", title = "(B)") +
  geom_text(data = filter(top15, name=="Last 5 years"),
            aes(label = package), 
            hjust = 0, 
            nudge_x = 0.1,
            size = 10) + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

```{r, fig.cap=glue("The above graph shows the total download counts from {start} to {end} of DoE packages."), fig.width = 6}
dl1plot + dl2plot
```

```{r text, cache=tocache}
url <- paste0(getOption("repos")["CRAN"], "web/packages/packages.rds")
db <- readRDS(url(url)) %>% 
  as.data.frame()

doe_db <- db %>% 
  filter(Package %in% doe_pkgs) %>% 
  mutate(Description = str_replace_all(Description, "\n", " "),
         Description = str_squish(Description),
         Title = str_replace_all(Title, "\n", " "))
```

```{r}
singularize2 <- function(x) {
    res <- singularize(x)
    res <- ifelse(res=="bia", "bias", res)
    ifelse(res=="baye", "bayes", res)
}
stop_words_rex <- c(paste0("^", stop_words$word, "$"), "^doi", "^[0-9.]+$")
get_ngram <- function(ngram, desc = c("Title", "Description")) {
  desc <- match.arg(desc)
  words <- paste0("word", 1:ngram)
  doe_db %>% 
    unnest_tokens(word, desc, token = "ngrams", n = ngram) %>% 
    separate(word, words, sep = " ") %>% 
    mutate(across(num_range("word", 1:ngram), singularize2)) %>% 
    distinct(!!!map(c("Package", words), as.name)) %>% 
    filter(if_all(num_range("word", 1:ngram), 
                  ~!str_detect(., paste0(stop_words_rex, collapse = "|")))) %>% 
    count(!!!map(words, as.name), sort = TRUE) %>% 
    mutate(word = do.call("paste", map(words, ~eval(parse(text = .x))))) %>% 
    select(word, n) 
}
```

## What are the common types of experimental designs?

## How do packages interplay with each other?

## Which packages are widely utilised?

# Software Design

# Propsects and Discussion

\newpage

```{r cran-titles}
url <- "http://cran.rstudio.com/web/packages/packages.rds"
db <- readRDS(url(url)) %>% 
  as.data.frame()
```

@@

```{r bigram-title, results="asis"}
btitle <- bigram_tab(doe_db, Title) %>% 
  filter(n > 3) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(booktabs = TRUE, linesep = "")

bdesc <- bigram_tab(doe_db, Description) %>% 
  filter(n > 4) %>% 
  mutate(word = paste(word1, word2)) %>% 
  select(word, n) %>% 
  kbl(booktabs = TRUE, linesep = "")

cat(c("\\begin{table}[h] \\centering ", 
      btitle,
    "\\hspace{1cm} \\centering ",
      bdesc,
    "\\caption{My tables} \\end{table}")) 
```

@@

@@

@@

### 2021-03-05 paper.tex
abstract + intro

\abstract{%
The critical role of data collection is well captured in the expression
``garbage in, garbage out'\,' -- in other words, if the collected data
is rubbish then no analysis, however complex it may be, can make
something out of it. The gold standard for data collection is through
well-designed experiments. Re-running an experiment is generally
expensive and in some cases difficult due to limited resources, The
design of experiment is important in getting a good value out of the
collected data. In this article I describe the current state of the
R-packages for the design of experiments through textual analysis and
download trends. I discuss also the software design of widely utilised R
packages in the field and conclude with discussion of some future
prospects for the field.
}

@@

@@

@@

\begin{Schunk}
\begin{table}

\caption{\label{tab:bigram-title}The bigram of the R-package titles as provided in the DESCRIPTION file in CRAN.}
\centering
\begin{tabular}[t]{lr}
\toprule
Bigram & Count\\
\midrule
optimal design & 10\\
experimental design & 8\\
clinical trial & 5\\
dose finding & 5\\
sequential design & 5\\
block design & 4\\
microarray experiment & 4\\
response surface & 4\\
\bottomrule
\end{tabular}
\end{table}

\end{Schunk}

\begin{Schunk}
\begin{table}

\caption{\label{tab:bigram-desc}The bigram of the R-package descriptions as provided in the DESCRIPTION file in CRAN.}
\centering
\begin{tabular}[t]{lr}
\toprule
Bigram & Count\\
\midrule
experimental design & 11\\
optimal design & 10\\
package provide & 7\\
response surface & 7\\
factorial design & 6\\
graphical user & 6\\
user interface & 6\\
block design & 5\\
contour plot & 5\\
design based & 5\\
effect model & 5\\
fractional factorial & 5\\
microarray experiment & 5\\
mixed effect & 5\\
provide function & 5\\
sample size & 5\\
sequential design & 5\\
\bottomrule
\end{tabular}
\end{table}

\end{Schunk}

\begin{Schunk}

\begin{center}\includegraphics{paper_files/figure-latex/download-hist-1} \end{center}

\end{Schunk}

\begin{Schunk}
\begin{Soutput}
#>  [1] "agricolae"   "AlgDesign"   "conf.design" "DiceDesign"  "DiceKriging"
#>  [6] "DoE.base"    "ez"          "FrF2"        "lhs"         "rsm"
\end{Soutput}
\end{Schunk}

\begin{Schunk}

\begin{center}\includegraphics{paper_files/figure-latex/download-barplot-1} \end{center}

\end{Schunk}

\begin{Schunk}

\begin{center}\includegraphics{paper_files/figure-latex/download-timeplot-1} \end{center}

\end{Schunk}

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

\texttt{test} \code{test}

@@

{\centering \includegraphics[width=2in]{/Users/etan0038/Dropbox/projects/paper-review-DoE-pkgs/Rlogo}

\caption[The logo of R]{The logo of R.}\label{fig:Rlogo}
\end{figure}
\end{Schunk}

>>>>

\abstract{%
Re-running an experiment is generally costly and in some cases
impossible due to limited resources, so the design of experiment plays a
critical role in increasing the quality of experimental data. In this
article I describe the current state of the R-packages for the design of
experiments through textual analysis and download trends. I discuss also
the software design of widely utilised R packages in the field of
experimental design and conclude with discussion of some future
prospects for the field.
}

@@

The critical role of data collection is well captured in the expression
``garbage in, garbage out'' -- in other words, if the collected data is
rubbish then no analysis, however complex it may be, can make something
out of it. Methods for data collection can be dichotomised by the type
of data collected -- namely, experimental or observational -- or
alternatively, categorised into experimental design or survey design.
Survey designs often, although not always, aim to collect observational
data whilst experimental designs are exclusively center on experimental
data. This dichotomisation, to a great extent, is seen in CRAN task
views where R-packages in experimental design are in
\ctv{ExperimentalDesign} and R-packages in survey designs are in
\ctv{OfficialStatistics}.

In the CRAN task view of \ctv{ExperimentalDesign}, there are 113 R
packages for experimental design and analysis of data from experiments,
henceforth referred to as ``DoE packages'' in this paper. The sheer
quantity and variation in the output experimental design in the
R-packages are arguably unmatched with any other programming languages,
e.g.~in Python \citep{python}, only a handful of libraries that generate
design of experiment exist (namely \texttt{pyDOE}, \texttt{pyDOE2},
\texttt{dexpy}, \texttt{experimenter} and \texttt{GPdoemd}) with limited
outputs.

@@

{\centering \includegraphics{paper_files/figure-latex/unnamed-chunk-1-1}

\caption[The above graph shows the total download counts from 2019-02-09 to 2021-02-07 of DoE packages]{The above graph shows the total download counts from 2019-02-09 to 2021-02-07 of DoE packages.}\label{fig:unnamed-chunk-1}
\end{figure}
\end{Schunk}

@@

\begin{table}[h] \centering  
\begin{tabular}[t]{lr}
\toprule
word & n\\
\midrule
optimal design & 10\\
experimental design & 8\\
clinical trial & 5\\
dose finding & 5\\
sequential design & 5\\
block design & 4\\
microarray experiment & 4\\
response surface & 4\\
\bottomrule
\end{tabular} \hspace{1cm} \centering  
\begin{tabular}[t]{lr}
\toprule
word & n\\
\midrule
experimental design & 11\\
optimal design & 10\\
package provide & 7\\
graphical user & 6\\
response surface & 6\\
user interface & 6\\
block design & 5\\
contour plot & 5\\
design based & 5\\
effect model & 5\\
factorial design & 5\\
microarray experiment & 5\\
mixed effect & 5\\
provide function & 5\\
sample size & 5\\
sequential design & 5\\
\bottomrule
\end{tabular} \caption{My tables} \end{table}

@@
